---
title: "Part 2 - chi-squared with lizards, tails"
author: "Renee LaManna"
date: "11/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(broom)
```

## Read in data 
```{r}
lizards <- read_csv(here("data", "lizards.csv"))
```

Get counts of lizards by species:
```{r}
lizards %>% 
  count(spp) %>% 
  arrange(-n)
#arrange in decreasing order by this variable n that contains the counts
```

## Make a subset

- UTST lizards
- Sites: IBPE (enclosure glassland). RABB (rabbit mesquite site), "EAST" (east tarbush site)
- Tail condition is recorded either as W = whole or B = broken

```{r}
utst <- lizards %>% 
  filter(spp == "UTST") %>% 
  filter(site %in% c("IBPE", "RABB", "EAST")) %>% 
# Use unique fxn to check things u think u are restricting in your filter are actually the things that show up in resulting data frame. e.g. in console: unique(utst$site)
  filter(tail %in% c("B", "W"))
```

## Find counts and proportions of lizrds with broken or whole tails at each site

```{r}
# count by site and tail. will tell us for each site how many lizards had recorded broken tail and how mnay had whole tail
utst %>% 
  count(site, tail)
```

```{r}
tail_counts <- utst %>% 
  janitor::tabyl(site, tail)

# proportions by row
tail_proportions <- tail_counts %>% 
  janitor::adorn_percentages() %>% 
  janitor::adorn_pct_formatting(digits = 2) %>% 
  janitor::adorn_ns()

# use adorn_ns, a package in janitor to include actual counts next to percentages
```

Is there a signif association between site and tail breakage?
- Do i think i have enough evidence to reject null that tail breakage is independent of site

## Convert to contingency table

- use fxn to convert column into row names so we can just have column of proportions/counts
- Use `column_to_rowname()`.

```{r}
lizard_ct <- tail_counts %>% 
  column_to_rownames(var = "site")
```

## Chi-square
Effect of site on tail breakage

```{r}
lizard_x2 <- chisq.test(lizard_ct)
lizard_x2

# get outputs in tidy format to better use for an inline code
lizard_tidy <- broom::tidy(lizard_x2)
```
This X-squared value of 13.064 (how diff values in our table our from expected null hypothesis values) with 2 degrees of freedom (rows-1 * columns-1) results in a p-value of 0.001456. Yes we have suffiecent evidence tio reject null of independence which means we retain alternative hypothesis that site and tail breakage is not independent. 

There is a signif association between site and lizard tail breaking.

AKA. Site and lizard tail breakage are not independent

There is a significant association (i.e. non-independence)
between site and lizard tail breakage ($\chi$^2^(`r lizard_tidy$parameter`) = `r lizard_tidy$statistic`)\, *p* = `r lizard_tidy$p.value`).
